# Data

## Description

## Missing value analysis

```{r}
install.packages(c(
  "tidyverse", "data.table", "lubridate", "janitor",
  "stringr", "arrow", "scales", "glue",
  "readr", "forcats", "plotly", "patchwork"
))
```

```{r}
install.packages("janitor")
```

```{r}
install.packages('R.utils')
```

```{r}
install.packages('skimr')
```

```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(janitor)
library(stringr)
library(scales)
library(glue)
library(readr)
library(forcats)
library(plotly)
library(R.utils)
library(skimr)
```

```{r}
download_imdb <- TRUE
imdb_dir <- "data/imdb"
dir.create(imdb_dir, recursive = TRUE, showWarnings = FALSE)

imdb_base <- "https://datasets.imdbws.com"

files <- c(
"title.basics.tsv.gz",
"title.ratings.tsv.gz",
"title.akas.tsv.gz"
)
if(download_imdb){
for(f in files){
url <- file.path(imdb_base, f)
dest <- file.path(imdb_dir, f)
if(!file.exists(dest)){
message("Downloading ", f)
download.file(url, dest, mode = "wb")
} else {
message(f, " already exists; skipping download.")
}
}
}

```

```{r}
basics <- fread(
  file.path(imdb_dir, "title.basics.tsv.gz"),
  na.strings = "\\N",
  sep = "\t",
  quote = "",
  showProgress = FALSE
)

ratings <- fread(
  file.path(imdb_dir, "title.ratings.tsv.gz"),
  na.strings = "\\N",
  sep = "\t",
  quote = "",
  showProgress = FALSE
)

akas <- fread(
  file.path(imdb_dir, "title.akas.tsv.gz"),
  na.strings = "\\N",
  sep = "\t",
  quote = "",
  showProgress = FALSE
)

```

```{r}
glimpse(basics)
glimpse(ratings)
glimpse(akas)

```

Keep only movies (titleType == "movie" and isAdult == 0).

Split genres into list columns.

Convert years to numeric and create release_decade.

Join IMDb ratings.

```{r}
movies <- basics |>
  filter(titleType == "movie") |>
  mutate(
    startYear = as.integer(startYear),
    endYear = as.integer(endYear),
    runtimeMinutes = as.integer(runtimeMinutes),
    isAdult = as.integer(isAdult)
  ) |>
  select(
    tconst, primaryTitle, originalTitle,
    startYear, runtimeMinutes, genres, isAdult
  ) |>
  filter(isAdult == 0)

```

Join Ratings

```{r}
movies <- movies |>
left_join(ratings, by = c("tconst" = "tconst")) |>
rename(imdb_rating = averageRating, imdb_votes = numVotes)
```

Parse genres into List

```{r}
movies <- movies |>
  mutate(
    genres_list = if_else(
      is.na(genres),
      list(NA_character_),
      str_split(genres, ",\\s*")  
    ),
    release_decade = case_when(
      !is.na(startYear) ~ paste0(floor(startYear/10) * 10, "s"),
      TRUE ~ NA_character_
    )
  )
```

Inspect missing data

```{r}
missing_summary <- movies |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "column", values_to = "missing")

missing_summary

```

Handling Missing Values

```{r}
movies_clean <- movies |>
  # Remove movies missing release year (very small fraction)
  filter(!is.na(startYear)) |>
  # Remove movies with no ratings (not useful for analysis)
  filter(!is.na(imdb_rating), !is.na(imdb_votes)) |>
  # Remove movies with runtime 0 or unrealistic (< 10 min)
  filter(!is.na(runtimeMinutes), runtimeMinutes >= 10) |>
  # Keep only movies with at least 50 votes (reduces noise)
  filter(imdb_votes >= 50)
```

```{r}
# ---- Final Cleaned Dataset Summary ----

skimr::skim(movies_clean)

clean_missing_summary <- movies_clean |> 
  summarise(across(everything(), ~sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "column", values_to = "missing")

clean_missing_summary
```

Ratingd Over Decade

```{r}
# ---- Plot: Average IMDb Rating Over Decades ----
movies_clean |>
  group_by(release_decade) |>
  summarise(mean_rating = mean(imdb_rating)) |>
  ggplot(aes(x = release_decade, y = mean_rating, group = 1)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average IMDb Rating by Release Decade",
    x = "Decade",
    y = "Avg Rating"
  ) +
  theme_minimal()

```

Number of Movies Released per Decade

```{r}
# ---- Plot: Movie Count by Decade ----
movies_clean |>
  count(release_decade) |>
  ggplot(aes(x = release_decade, y = n)) +
  geom_col() +
  labs(
    title = "Number of Movies Released by Decade",
    x = "Decade",
    y = "Count"
  ) +
  theme_minimal()

```

Rating vs Runtime (with smoothing)

```{r}
# ---- Plot: Rating vs Runtime ----
movies_clean |>
  ggplot(aes(x = runtimeMinutes, y = imdb_rating)) +
  geom_point(alpha = 0.1) +
  geom_smooth() +
  labs(
    title = "Relationship Between Movie Runtime and IMDb Rating",
    x = "Runtime (min)",
    y = "IMDb Rating"
  ) +
  theme_minimal()

```

Top Genres by Count

```{r}
# ---- Unnest genres ----
genre_counts <- movies_clean |>
  unnest(genres_list) |>
  filter(!is.na(genres_list)) |>
  count(genres_list, sort = TRUE)

# ---- Plot ----
genre_counts |> 
  slice_max(n, n = 15) |>
  ggplot(aes(x = reorder(genres_list, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most Common IMDb Movie Genres",
    x = "Genre",
    y = "Count"
  ) +
  theme_minimal()

```

Genre Vs Average Rating

```{r}
genre_rating <- movies_clean |>
  unnest(genres_list) |>
  filter(!is.na(genres_list)) |>
  group_by(genres_list) |>
  summarise(
    mean_rating = mean(imdb_rating),
    count = n()
  ) |>
  filter(count >= 200)   # avoid tiny-sample genres

genre_rating |>
  ggplot(aes(x = reorder(genres_list, mean_rating), y = mean_rating)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average IMDb Rating by Genre (≥ 200 movies)",
    x = "Genre",
    y = "Average Rating"
  ) +
  theme_minimal()

```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```
